---
title: "Mapping Species Occurrences"
output: pdf_document
---

## Mapping Species Occurrences

Maxine Cruz, Caelan Wilkie-Rogers, and Deanna Sunnergren (?)
Milkfli
Mapping Species Occurrences 
February 15th 

There are two big goals of today's R Notebook:

* Practicing the integration of GitHub into your workflow. 
* Diving deeper into using R to map species distribution.  

DUE BY NEXT MONDAY: a jpg image of your species distribution map.


Setup instructions for your reference: https://docs.google.com/document/d/1EO5AH3AczU-ruHLnvZ9Nz7yn8NaSfwgzHYyqKi4ZYXY/edit?usp=sharing


```{r}
# load packages:

library("spocc")
library("sp")
library("raster")
library("maptools")
library("rgdal")
library("dismo")
library("sf")
library("tidyverse")

```




### GitHub integration


Workflow:
* write some code (typically some sort of project-oriented goal)
* commit. 
* push.
* pull (if you need to get the latest pushed code from github)
* Use Slack to communicate! (e.g. "ok, I just pushed to github, go ahead and pull)

Check out commit-push-pull.jpg for a visual representation.

How to Commit (i.e. selecting files to sync to GitHub): 
* Save any files you've been editing toward your "commit" goal.
* In the upper-right window, open the Git tab. 
* Click the check boxes next to any files to be included in your commit.
* Click "Commit"
* A window will open. Write a description of your commit in the "Commit Message" box, and click "Commit".
* Close the resulting window. 

How to Push (i.e. sync the files to GitHub): 
* Immediately after you've committed, you'll want to push.  
* In the window that had opened from your commit, click Push (next to green up arrow)
* A window will pop up with a message, which you can close.
* IF you're asked for a password, highlight and run the code in github.R, and try pushing again.

How to Pull (i.e. get the latest changes in GitHub into your rstudio.cloud environment):
* In the Git tab, click the down-pointing green arrow.

BEWARE: Things can get really messy if you and your groupmates are simultaneously working on the same file(s), and committing-pushing-pulling. For now, it's best to communicate, and take turns. If you run into problems, don't fret...we've all been there. Reach out to us on Slack, and we can help things get squared away.





Practice:

Group member 1: 
* write code in the code chunk below to query milkweed species from GBIF
* Commit
* Push
* Tell your groupmates to pull in Slack

```{r}
# Query (retrieve data on) Asclepias speciosa

# Initially (without the limit), we will only get 500/9353 observations:

# Searched: gbif
# Occurrences - Found: 9,353, Returned: 500
# Search type: Scientific
#   gbif: Asclepias speciosa (500)

# But now that we know what the maximum number of observations is (9353),
# we can add the limit so that it includes all observations.

showy_milkweed <- occ(query = "Asclepias speciosa",
                      from = "gbif",
                      limit = 9400)

showy_milkweed

# This gives us:

# Searched: gbif
# Occurrences - Found: 9,353, Returned: 9,353
# Search type: Scientific
#   gbif: Asclepias speciosa (9353)

# We need this as a dataset though, so converting it into one would make life easier

showy_milkweed <- showy_milkweed$gbif$data$Asclepias_speciosa

showy_milkweed

```


Group member 2:
* write code in the code chunk below to query milkweed species from GBIF
* Commit
* Push
* Tell your groupmates to pull in Slack

```{r}



```


Group member 3:
* write code in the code chunk below to query milkweed species from GBIF
* Commit
* Push
* Tell your groupmates to pull in Slack

```{r}



```

Group member 4 (if applicable):
* write code in the code chunk below to query milkweed species from GBIF
* Commit
* Push
* Tell your groupmates to pull in Slack

```{r}



```


### Data cleaning


Just like in the last exercise, it's important to look for errors in the data.
Check here for a refresher: https://github.com/BiodiversityDataScienceCorp/more-querying-gbif-data/blob/main/more-querying-gbif.Rmd

Group Member 1:

Check for and filter out any data in which occurrenceStatus=="Absent"
* write code below
* Commit
* Push
* Tell your groupmates to pull in Slack

```{r}

# Here we use the subset function to return rows from 
# the original data set based on a condition we specify

absent <- subset(x = showy_milkweed, occurrenceStatus !="PRESENT")

# Now we use "anti_join" to redefine our data set by 
# subtracting the "absent" rows from it
# The anti_join function will look at two tables that are inputted,
# then it will create a new table of values based on the former table that are
# not present in the latter table (a.k.a. the PRESENT data in this case)

showy_milkweed <- anti_join(showy_milkweed, absent)

```


Group Member 2:

Check for and filter out any occurrences in which individualCount==0
* write code below
* Commit
* Push
* Tell your groupmates to pull in Slack

```{r}




```


Group Member 3:

Check for obvious errors in lat/long. Remove those points if necessary.
* write code below
* Commit
* Push
* Tell your groupmates to pull in Slack

```{r}




```

Group Member 4 (or 1):

Reduce data columns, and save to a CSV file
* write code below
* Commit
* Push
* Tell your groupmates to pull in Slack
NOTE: Depending upon the size of your .csv file, you may not want to commit and push it. Your groupmates can always re-run the code to generate their own local copy.


### Mapping

We did a little bit of mapping in the last assignment, when we were checking for erroneous points. Now let's try to make a more fine-tuned map of species occurrences, with the goal of creating ans image you could use in your final report.

There are numerous ways to create a map in R. We're going to look at one method here, but don't feel restricted if you want to try a different technique. Searching in Google is an effective way to learn about alternatives.

Let's look at an example using the Goldfinch data.

First we want to load the data set:

```{r}
goldfinches<-read_csv("goldfinchOr.csv")
goldfinches
```


Now let's determine the minimum and maximum latitudes and longitudes of the data set. This will help us define where to focus the map. We want our boundaries to be slightly larger than all points contained:

```{r}

# ceiling returns the smallest integer that is greater than or equal to a value
# max gives the largest value of the column data
# so if the highest value in goldfinches$latitude is 46.5, then its ceiling is 47

max.lat <- ceiling(max(goldfinches$latitude))

# floor and min work similarly (but opposite)
# if the lowest value of goldfinches$latitude is 43.84, then its floor is 43

min.lat <- floor(min(goldfinches$latitude))

# run same calculations for longitude

max.lon <- ceiling(max(goldfinches$longitude))
min.lon <- floor(min(goldfinches$longitude))



```


Now these values are ready to use in the map code. 

In the sequence below, I'll first indicate that I want to save the output as a jpg. Then 


```{r}
# indicates that I want to save the next output as "or.jpg"
jpeg(file="or.jpg")

# this loads spatial polygons
data(wrld_simpl)

# Plot the base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon), # sets upper/lower x
     ylim = c(min.lat, max.lat), # sets upper/lower y
     axes = TRUE, 
     col = "grey95",
     main="Goldfinch species occurrences in Oregon, 2020",  # a title
     sub="an informative caption" # a caption
     )

# Add the points for individual observation
points(x = goldfinches$longitude, 
       y = goldfinches$latitude, 
       col = "blue", 
       pch = 20, 
       cex = 0.75)
# And draw a little box around the graph
box()

# indicates I'm done with plotting and saving to jpg
dev.off()
```


Select a group member to create a map of your milkweed occurrence data. Follow the example above: first determine your min/max lat/lon, and then generate the map code. Save this as a jpg file. Then commit, and push to Github (and have your group members pull).

```{r}

######## Query (retrieve data on) Asclepias speciosa ########

# Initially (without the limit), we will only get 500/9353 observations:

# Searched: gbif
# Occurrences - Found: 9,353, Returned: 500
# Search type: Scientific
#   gbif: Asclepias speciosa (500)

# But now that we know what the maximum number of observations is (9353),
# we can add the limit so that it includes all observations.

showy_milkweed <- occ(query = "Asclepias speciosa",
                      from = "gbif",
                      limit = 9400)

showy_milkweed

# This gives us:

# Searched: gbif
# Occurrences - Found: 9,353, Returned: 9,353
# Search type: Scientific
#   gbif: Asclepias speciosa (9353)



######## Refine query results ########

# We only want the milkweed data

showy_milkweed <- showy_milkweed$gbif$data$Asclepias_speciosa



######## Cleaning the data: occurenceStatus ########

# occurrenceStatus is data on whether the species was present or absent at a location

# This site gives the description for each of the columns in the data:
# https://dwc.tdwg.org/terms/#dwc:occurrenceStatus (for occurrenceStatus)

# This function checks for different values under occurrenceStatus in the GBIF data

unique(showy_milkweed$occurrenceStatus)

# The result from running this gives:
# [1] PRESENT

# This suggests that A. speciosa was present at said locations in the data

# Since the function did not return "ABSENT" or any other values, 
# we can assume that there is no need to filter for only "PRESENT" values

# [IN THE CASE THAT "ABSENT" IS RETURNED AFTER RUNNING UNIQUE()]

# Here we can use the subset function to return rows from 
# the original data set based on a condition we specify
# In this case, the subset is for values that are not equal to "PRESENT"

# absent <- subset(x = showy_milkweed, occurrenceStatus !="PRESENT")

# Now we can use anti_join to redefine our data set by 
# subtracting the "ABSENT" rows from it

# The anti_join function will look at two tables that are inputted,
# then it will create a new table of values based on the former table that are
# not present in the latter table (a.k.a. the PRESENT data in this case)

# showy_milkweed <- anti_join(showy_milkweed, absent)

# An alternative method could be using filter()



######## Cleaning the data: individualCount ########

# individualCount is the number of individuals observed at the time of occurrence

# The same procedure used for occurrenceStatus can be done here

# First, use unique() to find the different values under this data column

unique(showy_milkweed$individualCount)

# This will return:
# [1] NA  1  2  4  6  3

# It might be good to filter out the NA, but I need to ask about that
# Under the assumption that we only need to filter cases where the
# value is 0, we can assume there is no need to filter individualCount

# In any case, we can have an alternate dataset with the NA omitted

# Check to see if NA values were removed

showy_milkweed_noNA <- showy_milkweed[!is.na(showy_milkweed$individualCount), ]

# This returns:
# [1] 1 2 4 6 3

# [IN THE CASE THAT 0 IS RETURNED AFTER RUNNING UNIQUE()]

# zero <- subset(x = showy_milkweed, individualCount == 0)

# showy_milkweed <- anti_join(showy_milkweed, zero)



######## FOR THE PURPOSES OF MAPPING ########

# If NA values occur in the latitude and longitude,
# it may interfere with centering the map later on
# We can create a separate data set where the NA values are omitted

showy_milkweed_noNA_coord <- showy_milkweed[!is.na(showy_milkweed$latitude), ]
showy_milkweed_noNA_coord <- showy_milkweed[!is.na(showy_milkweed$longitude), ]



######## Cleaning the data: errors in lat/long ########

# We can first plot a map to visually assess whether points need to be removed

world_map <- borders("world", colour="gray50", fill="gray50")

sample_map <- ggplot() + 
  coord_fixed() + 
  world_map +
  geom_point(data = showy_milkweed,
             aes(x = longitude, y = latitude),
             colour = "darkred", 
             size = 1.0) +
  theme_bw()

# This results in a map that shows that A. speciosa is primarily in N. America
# There are a few points in Europe and Africa, and it is possible that
# these occurrences are native or established
# But in our case, we are interested in the occurrences in the Americas
# So we can modify our data to showcase only points in America

# We can see that the points in the Americas are mostly between
# Latitude: 0 degrees and 100 degrees
# Longitude: -50 degrees and -150 degrees

# Therefore, we can filter the data to those within those boundaries

showy_milkweed <- showy_milkweed %>% 
  filter(latitude <= 100 & latitude >= 0) %>%
  filter(longitude <= -50 & longitude >= -150)

# Replot data on world map to ensure filtering was correct

sample_map_2 <- ggplot() + 
  coord_fixed() + 
  world_map +
  geom_point(data = showy_milkweed,
             aes(x = longitude, y = latitude),
             colour = "darkred", 
             size = 1.0) +
  theme_bw()



######## Reduce columns in data ########

# We do not need all the columns given by GBIF
# So we can keep what will or might be necessary

showy_milkweed <- select(showy_milkweed, 
                         c(name, latitude, longitude,
                           scientificName, publishingCountry, stateProvince,
                           year, month, day, eventDate, individualCount))

showy_milkweed_noNA <- select(showy_milkweed_noNA, 
                         c(name, latitude, longitude,
                           scientificName, publishingCountry, stateProvince,
                           year, month, day, eventDate, individualCount))

showy_milkweed_noNA_coord <- select(showy_milkweed_noNA_coord, 
                         c(name, latitude, longitude,
                           scientificName, publishingCountry, stateProvince,
                           year, month, day, eventDate, individualCount))


######## Save data as .csv file ########

write_csv(showy_milkweed, "showy_milkweed.csv")
write_csv(showy_milkweed_noNA, "showy_milkweed_noNA.csv")
write_csv(showy_milkweed_noNA_coord, "showy_milkweed_noNA_coord.csv")



######## Load .csv data file for mapping ########

csv_showy_milkweed_noNA_coord <- read_csv("showy_milkweed_noNA_coord.csv")



######## Determine max and min latitudes and longitudes to center map ########

# We do not want an off-centered map
# So it would be good to find the boundaries of where the data are
# In doing so, we can plot a map that is focused in on that region

# ceiling() gives the smallest integer that is greater than or equal to a value
# max() gives the largest value of the column data
# So if the highest value in dataset$latitude is 46.5, then its ceiling is 47

max.lat <- ceiling(max(csv_showy_milkweed_noNA_coord$latitude))

# floor() and min() work similarly (but opposite)
# If the lowest value of dataset$latitude is 43.84, then its floor is 43

min.lat <- floor(min(csv_showy_milkweed_noNA_coord$latitude))

# Run the same calculations for longitude

max.lon <- ceiling(max(csv_showy_milkweed_noNA_coord$longitude))
min.lon <- floor(min(csv_showy_milkweed_noNA_coord$longitude))



######## Mapping points for A. speciosa ########

# This indicates that we want to save the next output as "or.jpg"

jpeg(file = "or.jpg")

# This loads spatial polygons
# Spatial polygons are shapes/polygons that represent a geographic location

data(wrld_simpl)

# Plot the base map

plot(wrld_simpl, 
     xlim = c(min.lon, max.lon), # Sets upper/lower x
     ylim = c(min.lat, max.lat), # Sets upper/lower y
     axes = TRUE, 
     col = "grey95",
     main = "Asclepias speciosas occurrences in the Americas")
     
# Add the points for individual observations

points(x = csv_showy_milkweed_noNA_coord$longitude, 
       y = csv_showy_milkweed_noNA_coord$latitude, 
       col = "blue", 
       pch = 20, 
       cex = 0.75)

# And draw a little box around the graph

box()

# This indicates we are done with plotting and saving to .jpg

dev.off()

```


Download your map jpg, and submit it to the assignment 5 folder. Name it in the format of "Assignment5_GroupName.jpg".  

Turn in the image to the Assignment 5 folder here: https://drive.google.com/drive/folders/1pJW4cqtamx_AvldgajF9SkhY9uxz32EW?usp=sharing
